Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-07-04 manuscript/manuscript.Rmd
reorganization of section one.

The now extensive corpus of experimental archaeology have witnessed the growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

Controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms another major issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone tool replication, it can be projected into the debate over the use of machines in knapping, a research design that has received increasing attention in the past decades []. Machine knapping is a typical design with high internal validity but low external validity, which has been proved to provides critical insights into lithic fracture mechanics by identifying potential causal variables at the level of individual stone artifact such as determining which angle of blow or how much force of blow will produce the maximal amount of blade area. All the variables of interest are easy to measure, quantify, and control in a machine knapping setting. Nevertheless, being easy to control is not always a virtue as it essentially eliminates the potential interactions between variables operable in the past and thereby providing misleading results when answering archaeological questions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks or video instruction in in a series of recently published experimental studies aiming at understanding the transmission of lithic technologies. As a rule of thumb, external validity should be given more weight in the design when the research focuses on the behaviors of the users of artifacts while internal validity matters more when it comes to the properties of artifacts themselves.

In the past decades, actualistic experiments becomes more common [@liu2022]. Variability as revealed in experiment is crucial and cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023]. It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020]. how decontexualized law-like statements interact in a particular context.

Experimental archaeology is based on the concept of analogy (i.e., the past is at least partially similar to the present in some aspects). It is acknowledged that the validity of this type of analogical inference has long been a subject of debate in archaeology [@wylie1985; @chapman2016], and a comprehensive review of it is beyond the scope of this essay. For instance, presumably in the paleolithic period the development of stone knapping skills mostly happened during childhood and these children grew up in an environment surrounded by habitual stone tool users and makers. In this case, what can knapping teaching experiments involving modern adults who have zero exposure to stone tools inform us about the past learning behavior? It is important to clarify that no experimental project ever intends or claims to provide the perfect reconstruction of the past but rather aims at identifying variables relevant to the question of interest [@stoutSkillLearningHuman2015], which can be often ignored through pure deductive reasoning. In the end, all experiments are wrong, but some are useful, and we need more of them.

Again, actualistic expeirment as a huestric for hypothesis generation as well as a low-cost , echoing the spirit stated by Lin et al. [-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".

@@

Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020] and museum collection studies [@timbrell2022] of stone artifacts.

@@

>>>>

In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at behavioral level in the explanation of material culture variation [@eren2016; @lin2018; @rezek2020; @outram2008; @reynolds1999]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding the fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights on our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

Trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped identifying multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapbale of inferring how decontexualized law-like statements interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acqusition. After all, these experimental results can only be as robust as its experimental settings.

More importantly, controlled experiments without randomization are not enough to infer causal mechanism, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past are derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the golden standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

On the other hand, actualistic experiments pay more attention to how experimental results can be generated to archaeological samples. Unlike controlled experiment, variability could be easily observed in actualistic experiment by design, and it cannot be simply replaced by ethnographic records because the many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with ground stone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data has also been greatly boosted in recent years [@cunningham2021; @hernan2023]. It also echoes the recent calls for moving beyond the by providing detailed contextual information of the experiment [@holleman2020]. Lastly, actualistic experiment can act as a hurestric for hypothesis generation, echoing the spirit stated by Lin et al. [-@lin2018: 680-681] as well as Ingersoll and MacDonald [-@ingersoll1977] arguing that the the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction".

@@

Emphasizing variability at its core, the Triple P conceptual framework inherently adopts an collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts.

@@

# Conclusion

The Triple P framework is theoretically compatible with
